---
title: "Project Draft"
author: "David Tessier"
date: "11/09/2021"
output: html_document
---
In this draft, our figures show the EPA and WPA per first down for QB drop backs vs. run plays of each team in the years 2015-2020. This helps to show how effective a team was compared to the rest of the league in either type of play attempt, and we can further show this and other models for individual seasons, weeks, and games. This will assist us in exploring our thesis; we will analyze the relationship and effectiveness of Qb drop backs and run plays on first and 10, proving or disproving that teams have higher success on QB drop backs. 

### Required Packages
```{r}
#install.packages("tidyverse", type = "binary")
#install.packages("ggrepel", type = "binary")
#install.packages("ggimage", type = "binary")
#install.packages("nflfastR", type = "binary")
#install.packages("gsisdecoder")
#install.packages("ggplot2")

library(tidyverse, warn.conflicts = FALSE)
library(ggrepel, warn.conflicts = FALSE)
library(ggimage, warn.conflicts = FALSE)
library(nflfastR, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
library(nflreadr, warn.conflicts = FALSE)
library(nflscrapR, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)

# makes R prefer not to display numbers in scientific notation
options(scipen = 9999)

```

```{r}
# Loading all regular season games from 2015 - 2020
data1 <- load_pbp(2015:2020) %>% 
  filter(season_type == "REG")

team_abbr <- c("ARI", "ATL", "BAL", "BUF", "CAR", "CHI", "CIN", "CLE",  "DAL", "DEN", "DET", "GB", "HOU", "IND", "JAX", "KC", "LA", "LAC", "LV", "MIA", "MIN", "NE", "NYG", "NYJ", "NO", "PHI", "PIT", "TEN", "SEA", "SF", "TB", "WAS")

################################################################################
#   QB Drop back plays on 1st and 10 with win probability between 20 and 80% ,
#   excluding the final 2 minutes of the either half, excluding plays with a 
#   line of scrimmage between either half's 20 yard line and goal line.
fd_dropbacks1 <- data1 %>%
  filter(down==1 & ydstogo==10 & yardline_100<=80 & yardline_100>=20 & qb_dropback==1 & wp>= .20 & wp<= .80 & half_seconds_remaining>120 & !is.na(epa)) %>%
  select(posteam, season, week, play_id, down, ydstogo, epa, wpa, success, yards_gained)

################################################################################
#   Run plays on 1st and 10 under same conditions as fd_dropbacks1
fd_runs1 <- data1 %>%
  filter(down==1 & ydstogo==10 & yardline_100<=80 & yardline_100>=20 & 
         qb_dropback==0 & play_type == "run" & wp>= .20 & wp<= .80 & 
         half_seconds_remaining>120 & !is.na(epa)) %>%
  select(posteam, season, week, play_id, down, ydstogo, epa, wpa, success, yards_gained)
```

### Evaluate by week
```{r}
fd_dropbacks_by_week <- fd_dropbacks1 %>%
  group_by(posteam, season, week)%>% 
  summarize(
    epa_fd_dropback = mean(epa),
    wpa_fd_dropback = mean(wpa),
    success_rate_fd_dropback = mean(success),
    yards_per_fd_dropback_play = mean(yards_gained), 
    fd_dropback_plays = n()
    )  
#fd_dropbacks_by_week

fd_runs_by_week <- fd_runs1 %>%
  group_by(posteam,season,week) %>%
  summarize(
    epa_fd_run = mean(epa),
    wpa_fd_run = mean(wpa),
    success_rate_fd_run = mean(success),
    yards_per_fd_run_play = mean(yards_gained),
    fd_run_plays = n()
  ) %>%
  ungroup()
#fd_runs_by_week

fd_plays_by_week <- fd_runs_by_week%>%
  left_join(fd_dropbacks_by_week , by = c('posteam' = 'posteam','season' = 'season', 'week' = 'week')) %>%
  left_join(teams_colors_logos , by = c('posteam' = 'team_abbr'))%>%
  mutate(tot_fd_plays = fd_run_plays + fd_dropback_plays,
         prop_fd_run_plays = fd_run_plays/tot_fd_plays,
         prop_fd_QBD_plays = fd_dropback_plays/tot_fd_plays)
fd_plays_by_week
```
### Evaluate by season
```{r}
### Data for QB drop backs on 1st and 10, grouped by season
fd_dropbacks_by_year <- fd_dropbacks1 %>%
  group_by(posteam, season)%>% 
  summarize(
    epa_fd_dropback = mean(epa),
    wpa_fd_dropback = mean(wpa),
    success_rate_fd_dropback = mean(success),
    yards_per_fd_dropback_play = mean(yards_gained), 
    fd_dropback_plays = n()
    ) %>%
  ungroup()
fd_dropbacks_by_year

### Data for runs on 1st and 10, grouped by season
fd_runs_by_year <- fd_runs1 %>%
  group_by(posteam, season)%>% 
  summarize(
    epa_fd_run = mean(epa),
    wpa_fd_run = mean(wpa),
    success_rate_fd_run = mean(success),
    yards_per_fd_run_play = mean(yards_gained),
    fd_run_plays = n()
    ) %>%
  ungroup()
fd_runs_by_year

### Data for both QB dropbacks and run plays on 1st and 10, grouped by season
fd_plays_by_year <- fd_runs_by_year%>%
  left_join(fd_dropbacks_by_year , by = c('posteam' = 'posteam','season' = 'season')) %>%
  left_join(teams_colors_logos , by = c('posteam' = 'team_abbr'))%>%
  mutate(tot_fd_plays = fd_run_plays + fd_dropback_plays,
         prop_fd_run_plays = fd_run_plays/tot_fd_plays,
         prop_fd_QBD_plays = fd_dropback_plays/tot_fd_plays)
fd_plays_by_year
```

### 2015
```{r}
fd_plays_2015 <- fd_plays_by_year %>%
  filter(season == 2015) 

run_epa_2015 <- fd_plays_2015 %>%
  ggplot(aes(x = prop_fd_run_plays, y = epa_fd_run)) +
  #horizontal line with mean epa on runs
  geom_hline(yintercept = mean(fd_plays_2015$epa_fd_run),
            color = "red", linetype = "dashed", alpha=0.5) +
  #vertical line with mean proportion of play calls being a run
  geom_vline(xintercept = mean(fd_plays_2015$prop_fd_run_plays),
             color = "red", linetype = "dashed", alpha=0.5) +
  #add points for the teams with the right logos
  geom_image(aes(image = team_logo_espn), size = fd_plays_2015$success_rate_fd_dropback / 9, asp = 16 / 9) +
  #add names using ggrepel, which tries to make them not overlap
  #geom_text_repel(aes(label=posteam)) +
  stat_smooth(geom = 'line', alpha = 0.5, se = FALSE, method = 'lm') +
  labs(x = "Prop. of Run Playcalls", 
       y = "EPA per Run on 1st & 10",
       title = "Run Play Efficiency on 1st Down, 2015",
       caption = "Data: @nflfastR") +
  theme_bw() +
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold")) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))
run_epa_2015

QBD_epa_2015 <- fd_plays_2015 %>%
  ggplot(aes(x = prop_fd_QBD_plays, y = epa_fd_dropback)) +
  #horizontal line with mean epa on dropbacks
  geom_hline(yintercept = mean(fd_plays_2015$epa_fd_dropback),
            color = "red", linetype = "dashed", alpha=0.5) +
  #vertical line with mean proportion of play calls being a QBD
  geom_vline(xintercept = mean(fd_plays_2015$prop_fd_QBD_plays),
             color = "red", linetype = "dashed", alpha=0.5) +
  geom_image(aes(image = team_logo_espn), size = fd_plays_2015$success_rate_fd_dropback / 9, asp = 16 / 9) +
  stat_smooth(geom = 'line', alpha = 0.5, se = FALSE, method = 'lm') +
  labs(x = "Prop. of QBD Playcalls", 
       y = "EPA per QBD on 1st & 10",
       title = "QB Drop Back Efficiency on 1st Down, 2015",
       caption = "Data: @nflfastR") +
  theme_bw() +
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold")) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))
QBD_epa_2015
```

### 2015 - 2020
```{r}
## Data for both QB dropbacks and run plays on 1st and 10 from 2015-2020
fd_plays_all_yrs <- fd_plays_by_year%>%
  group_by(posteam)%>%
  summarize(
    epa_fd_run = mean(epa_fd_run),
    wpa_fd_run = mean(wpa_fd_run),
    success_rate_fd_run = mean(success_rate_fd_run),
    yards_per_fd_run_play = mean(yards_per_fd_run_play),
    fd_run_plays = sum(fd_run_plays),
    epa_fd_dropback = mean(epa_fd_dropback),
    wpa_fd_dropback = mean(wpa_fd_dropback),
    success_rate_fd_dropback = mean(success_rate_fd_dropback),
    yards_per_fd_dropback_play = mean(yards_per_fd_dropback_play), 
    fd_dropback_plays = sum(fd_dropback_plays),
    tot_fd_plays = fd_run_plays + fd_dropback_plays,
         prop_fd_run_plays = fd_run_plays/tot_fd_plays,
         prop_fd_QBD_plays = fd_dropback_plays/tot_fd_plays
    
    ) %>%
  ungroup()

## Run epa graph from 2015-2020
run_epa_all_years <- fd_plays_all_yrs %>%
  left_join(teams_colors_logos , by = c('posteam' = 'team_abbr')) %>%
  ggplot(aes(x = prop_fd_run_plays, y = epa_fd_run)) +
  #horizontal line with mean epa on runs
  geom_hline(yintercept = mean(fd_plays_all_yrs$epa_fd_run),
            color = "red", linetype = "dashed", alpha=0.5) +
  #vertical line with mean proportion of play calls being a run
  geom_vline(xintercept = mean(fd_plays_all_yrs$prop_fd_run_plays),
             color = "red", linetype = "dashed", alpha=0.5) +
  #add points for the teams with the right logos
  geom_image(aes(image = team_logo_espn), size = fd_plays_all_yrs$success_rate_fd_run / 7, asp = 16 / 9) +
  #add names using ggrepel, which tries to make them not overlap
  #geom_text_repel(aes(label=posteam)) +
  stat_smooth(geom = 'line', alpha = 0.5, se = FALSE, method = 'lm') +
  labs(x = "Prop. of Run Playcalls", 
       y = "EPA per Run on 1st & 10",
       title = "Run Play Efficiency on 1st Down, 2015-2020",
       caption = "Data: @nflfastR") +
  theme_bw() +
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold")) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))
run_epa_all_years

## QBD epa graph from 2015-2020
QBD_epa_all_years <- fd_plays_all_yrs %>%
  left_join(teams_colors_logos , by = c('posteam' = 'team_abbr')) %>%
  ggplot(aes(x = prop_fd_QBD_plays, y = epa_fd_dropback)) +
  #horizontal line with mean epa on QBD
  geom_hline(yintercept = mean(fd_plays_all_yrs$epa_fd_dropback),
            color = "red", linetype = "dashed", alpha=0.5) +
  #vertical line with mean proportion of play calls being a QBD
  geom_vline(xintercept = mean(fd_plays_all_yrs$prop_fd_QBD_plays),
             color = "red", linetype = "dashed", alpha=0.5) +
  #add points for the teams with the right logos
  geom_image(aes(image = team_logo_espn), size = fd_plays_all_yrs$success_rate_fd_dropback / 9, asp = 16 / 9) +
  #add names using ggrepel, which tries to make them not overlap
  #geom_text_repel(aes(label=posteam)) +
  stat_smooth(geom = 'line', alpha = 0.5, se = FALSE, method = 'lm') +
  labs(x = "Prop. of QBD Playcalls", 
       y = "EPA per QBD on 1st & 10",
       title = "QBD Efficiency on 1st Down, 2015-2020",
       caption = "Data: @nflfastR") +
  theme_bw() +
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold")) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))
QBD_epa_all_years

```


### Evaluate all "1st and 10"s that began with a run play and combine this new data set with QB Dropbacks on 1st and 10s
```{r}







first_down_data %>%
  ggplot(aes(x = wpa_run, y = wpa_dropback)) +
  #horizontal line with mean win probability added on dropbacks
  geom_hline(yintercept = mean(first_down_data$wpa_dropback),
            color = "red", linetype = "dashed", alpha=0.5) +
  #vertical line with mean win probability added on runs
  geom_vline(xintercept = mean(first_down_data$wpa_run),
             color = "red", linetype = "dashed", alpha=0.5) +
  #add points for the teams with the right colors
  #cex controls point size and alpha the transparency (alpha = 1 is normal)
  geom_point(color = first_down_data$team_color, cex= (first_down_data$success_rate_dropback / first_down_data$success_rate_run) * 2, alpha = .6) +
  #add names using ggrepel, which tries to make them not overlap
  geom_text_repel(aes(label=posteam)) +
  stat_smooth(geom = 'line', alpha = 0.5, se = FALSE, method = 'lm') +
  labs(x = "Win Probability Added per Run", 
       y = "Win Probability Added per QB Dropback",
       title = "WPA per QB Dropback vs Run Play on 1st Down, 2015-2020",
       caption = "Data: @nflfastR") +
  theme_bw() +
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold")) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))

```
```{r}
first_down_data %>%
  ggplot(aes(x = success_rate_run, y = success_rate_dropback)) +
  #horizontal line with mean success rate on dropbacks
  geom_hline(yintercept = mean(first_down_data$success_rate_dropback),
            color = "red", linetype = "dashed", alpha=0.5) +
  #vertical line with mean success rate on runs
  geom_vline(xintercept = mean(first_down_data$success_rate_run),
             color = "red", linetype = "dashed", alpha=0.5) +
  #add points for the teams with the right colors
  #cex controls point size and alpha the transparency (alpha = 1 is normal)
  geom_point(color = first_down_data$team_color, cex= (first_down_data$success_rate_dropback / first_down_data$success_rate_run) * 2, alpha = .6) +
  #add names using ggrepel, which tries to make them not overlap
  geom_text_repel(aes(label=posteam)) +
  stat_smooth(geom = 'line', alpha = 0.5, se = FALSE, method = 'lm') +
  labs(x = "Success Rate per Run", 
       y = "Success Rate per QB Dropback",
       title = "Success Rate
       per QB Dropback vs Run Play on 1st Down, 2015-2020",
       caption = "Data: @nflfastR") +
  theme_bw() +
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold")) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))
```

```{r}
league_run_averages <- data1 %>%
  filter(down==1:3 & ydstogo>=5 & ydstogo<=10 & yardline_100<=80 & yardline_100>=20 & qb_scramble==0 & play_type == "run" & wp>=.20 & wp <= .80
         & half_seconds_remaining>120 & !is.na(epa)) %>%
  group_by(season) %>%
  summarize(
    league_epa_run = mean(epa),
    league_wpa_run = mean(wpa),
    league_success_rate_run = mean(success),
    league_yards_per_run_play = mean(yards_gained),
    league_run_plays = n()
  ) %>%
  ungroup()
league_run_averages

league_QBD_averages <- data1 %>%
  filter(down==1:3 & ydstogo>=5 & ydstogo<=10 & yardline_100<=80 & yardline_100>=20 & play_type == "pass" & wp>=.20 & wp <= .80
         & half_seconds_remaining>120 & !is.na(epa)) %>%
  group_by(season) %>%
  summarize(
    league_epa_QBD = mean(epa),
    league_wpa_QBD = mean(wpa),
    league_success_rate_QBD = mean(success),
    league_yards_per_QBD_play = mean(yards_gained),
    league_QBD_plays = n()
  ) %>%
  ungroup()
league_QBD_averages

data2 <- first_down_data %>%
  left_join(league_run_averages, by = "season") %>%
  left_join(league_QBD_averages, by = "season")
```





















